### 2015-01-08 prelim/oral_prelim.tex
prelim updates

\begin{frame}{Motivation}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item Average behavior is important to understand, but it does not paint the whole picture
    \begin{itemize}
      \item e.g. When constructing river levees, engineers need to be able to estimate a 100-year or 1000-year flood levels
      \item e.g. Probability of exceeding a certain threshold level
    \end{itemize}
    \item Spatial methods borrow information across space to estimate spatial correlation and make predictions by Kriging at unknown locations
    \item Want to explore similar methods for extremes
  \end{itemize}
\end{frame}

@@

\begin{frame}{Motivation}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Ozone compliance for Clean Air Act (EPA)
    \begin{itemize}
      \item Annual fourth-highest daily maximum 8-hour concentration, averaged over 3 years, not to exceed 75 ppb
      \item Annual fourth-highest is the 99th percentile for the year
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Defining extremes}
\begin{columns}[c]
\column{.45 \linewidth}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Key in extreme value analysis is to define extremes
    \item Typically done in one of two ways
    \begin{itemize}
      \item Block maxima
      \begin{itemize}
        \item Red dots
      \end{itemize}
      \item Over threshold
      \begin{itemize}
        \item Values over threshold considered extreme
      \end{itemize}
    \end{itemize}
  \end{itemize}

  \column{.55\linewidth}
  \begin{figure}
    \includegraphics[width=1\linewidth, trim=0 0.5in 0 1in]{./plots/define_extreme.pdf}
    \caption{Monthly maximums recorded over a three years}
    \end{figure}
\end{columns}
\end{frame}

\begin{frame}{Standard non-spatial analysis: Block maxima}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Asymptotic result (Falk et al., 2011)
    \begin{itemize}
      \item Let $X_1, \ldots, X_n$ be i.i.d.
      \item Consider $M_n = \max(X_1, \ldots, X_n)$
      \item If there exist normalizing sequences $a_n > 0$ and $b_n \in \calR$ such that
      \begin{align*}
        a_n^{-1} (M_n - b_n) \converged G(z)
      \end{align*}
      then $G(z)$ follows a generalized extreme value distribution
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Standard non-spatial analysis: Block maxima}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Generalized extreme value distribution has three parameters:
    \begin{itemize}
      \item $\mu \in \calR$ is a location parameter
      \item $\sigma > 0$ is a scale parameter
      \item $\xi \in \calR$ is a shape parameter
      \begin{itemize}
        \item Unbounded above if $\xi \ge 0$
        \item Bouded above by $(\mu - \sigma) / \xi$ when $\xi < 0$
      \end{itemize}
    \end{itemize}
    \item Challenge:
    \begin{itemize}
      \item Lose information by only considering maximum in a block
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Standard non-spatial analysis: Block maxima}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Generalized extreme value distribution
    \begin{align*}
      G(y) = \Pr(Y < y) = \left\{  \begin{array}{ll}
        \exp\left\{ -\left[ 1 + \xi \left( \frac{ y - \mu }{ \sigma } \right) \right]^{ -1 / \xi} \right\} & \quad \xi \neq 0 \\[0.5em]
        \exp \left\{ -\exp \left( - \frac{ y - \mu }{ \sigma} \right) \right\} & \quad \xi = 0
      \end{array}\right.
    \end{align*}
    \item Standardized distribution is unit Fr\'{e}chet or GEV(1, 1, 1)
    \begin{align*}
      \Pr(Z < z) = exp(-z^{-1})
    \end{align*}
  \end{itemize}
\end{frame}

\begin{frame}{Standard non-spatial analysis: Peaks over threshold}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Asymptotic result (Falk et al., 2011)
    \begin{itemize}
      \item Let $X_1, \ldots, X_n \sim F$
      \item If there exist normalizing sequences $a_n > 0$ and $b_n \in \calR$ with
      \begin{align*}
        \lim_{n \rightarrow \infty} 1 - F(b_n) = 0 \qquad \text{and} \qquad \lim_{ n \rightarrow \infty} \frac{ 1 - F(b_{n+1})}{1 - F(b_n)} = 1
      \end{align*}
      where $b_n$ is a sequence of thresholds
      \item If for any $x \ge 0$,
      \begin{align*}
        1 - \frac{ 1 - F(a_n x + b_n) }{ 1 - F(b_n) } \converged H(x),
      \end{align*}
      then $H(x)$ follows a generalized Pareto distribution
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Standard non-spatial analysis: Peaks over threshold}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Generalized Pareto distribution has two parameters:
    \begin{itemize}
      \item $\sigma > 0$ is a scale parameter
      \item $\xi \in \calR$ is a shape parameter
      \begin{itemize}
        \item Unbounded above if $\xi \ge 0$
        \item Bounded above by $(\mu - \sigma) / \xi$ when $\xi < 0$
      \end{itemize}
    \end{itemize}
    \item Challenges:
    \begin{itemize}
      \item Sensitive to threshold selection
      \item Temporal dependence between observations (e.g. flood levels don't dissipate overnight)
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Standard non-spatial analysis: Peaks over threshold}
  \begin{itemize}  \setlength{\itemsep}{0.5em}
    \item Select a threshold, $T$, and use the Generalized Pareto distribution to model the exceedances
    \begin{align*}
      H(y) = P(Y < y) = \left\{ \begin{array}{ll}
        1 - \left[1 - \xi \left( \frac{ y - T }{ \sigma } \right) \right]^{-1 / \xi} & \quad \xi \neq 0 \\[0.5em]
        1 - \exp \left\{ \frac{ y - T }{ \sigma} \right\} & \quad \xi = 0
      \end{array}\right.
    \end{align*}
    \item Related to GEV distribution through
    \begin{align*}
      H(y) = 1 + \log[G(y)], \quad y \ge T
    \end{align*}
  \end{itemize}
\end{frame}

\begin{frame}{Max-stable processes}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item For a spatial analysis, max-stable processes give an appropriate limiting distribution (Cooley et al., 2012)
    \begin{itemize}
      \item Consider a spatial process $x_t(\bs)$, $t = 1, \ldots, T$
      \item Let $M_T(\bs) = \left\{ \bigvee_{t=1}^T x_t(\bs_1), \ldots, \bigvee_{t=1}^T x_t(\bs_n) \right\}$
      \item If there exists normalizing sequences $a_T(\bs)$ and $b_T(\bs)$ such
      that for all sites, $\bs_i, i = 1, \ldots, d$,
      \begin{align*}
        a_T^{-1}(\bs) \left\{ M_T(\bs) - b_T(\bs) \right\} \converged Y(\bs)
      \end{align*}
      which has a non-degenerate distribution, then $Y(\bs)$ is a max-stable process
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Multivariate representations}
  \begin{itemize}
    \item Marginally at each site, observations follow a generalized extreme value distribution
    \item For a finite collection of sites
    \begin{itemize}
      \item The multivariate representation for the GEV is
      \begin{align*}
        \Pr(\bZ \le \bz)  &= G^*(\bz) = \exp(-V(\bz))\\
                V(\bs)    &= d \int_{\Delta_d} \bigvee_{i = 1}^d \frac{w_i}{z_i} H(\ddd w)
      \end{align*}
      where
      \begin{itemize}
        \item $\Delta_d = \{ \bw \in \calR^d_+ \mid w_1 + \cdots + w_d = 1\}$
        \item $H$ is a probability measure on $\Delta_d$
        \item $\int_{\Delta_d}w_i H(\ddd w) = 1 / d$ for $i = 1, \ldots, d$.
      \end{itemize}
    \end{itemize}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Multivariate peaks over threshold}
  \begin{itemize}
    \item Not a lot of existing methods
    \item Often use max-stable methods due to the relationship between GEV and GPD
    \item Joint distribution function given by Falk et al. (2011)
    \begin{align*}
      H(\bz) = 1 - V(\bz)
    \end{align*}
    where $V(\bz)$ is defined as in the GEV
  \end{itemize}
\end{frame}

\begin{frame}{Extremal dependence: $\chi$ statistic}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Correlation is the most common measure of dependence, but it is irrelevant for extreme value analysis
    \item Extreme value analysis focuses on the $\chi$ statistic, a measure of extremal dependence
    \item Specifically, we focus on $\chi(h)$ for the upper tail given by
    \begin{align*}
      \chi(h) = \lim_{c \rightarrow \infty}\Pr(Y(\bs) > c \mid Y(\bt) > c)
    \end{align*}
    where $h = ||\bs - \bt||$
    \item If $ \chi(h) = 0$, then observations are asymptotically independent at distance $h$
  \end{itemize}
\end{frame}

\begin{frame}{Existing challenges}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Multivariate max-stable and GPD models have nice features, but they are
    \begin{itemize}
      \item Computationally challenging (Falk et al., 2011)
      \begin{itemize}
        \item Asymmetric logistic has $2^{n-1}(n + 2) - (2n + 1)$ free parameters
      \end{itemize}
      \item Joint distribution only available in low dimension
    \end{itemize}
    \item Some recent approaches
    \begin{itemize}
      \item Bayesian hierarchical model (Reich and Shaby, 2012)
      \item Pairwise likelihood approach (Huser and Davison, 2014)
    \end{itemize}
    \item Many opportunities to explore new methods
  \end{itemize}
\end{frame}

\begin{frame}{Three principal contributions}
  \begin{enumerate}[1] \setlength{\itemsep}{0.5em}
    \item A spatio-temporal model with flexible tails, asymptotic spatial dependence, and computation on the order of Gaussian models for large space-time datasets
    \item Predicting rare binary events with a spatially dependent generalized extreme-value link function
    \item A Bayesian hierarchical model to allow for non-stationary covariance in extreme value models
  \end{enumerate}
\end{frame}

\begin{frame}{Spatiotemporal modeling for extreme values}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Model to analysis spatiotemporal extreme values
    \item Model objectives
    \begin{itemize}
      \item Has marginal distribution with a flexible tail
      \begin{itemize}
        \item Allow for asymmetric distributions
        \item Allow for heavy tails
      \end{itemize}
      \item Has asymptotic spatial dependence
      \item Has computation on the order of Gaussian models for large space-time datasets
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Gaussian spatial model}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item In geostatistics $Y(\bs)$ are often modeled using a Gaussian process with mean function $\mu(\bs)$ and covariance function $\rho(h)$.
    \item Model properties
    \begin{itemize}
      \item Nice computing properties (closed-form likelihood)
      \item For a Gaussian spatial model $\lim_{c \rightarrow \infty} \chi(h) = 0$ regardless of the strength of the correlation in the bulk of the distribution
      \item Tail is not flexible
      \begin{itemize}
        \item Light-tailed
        \item Symmetric
      \end{itemize}
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Spatial skew-$t$ distribution}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Assume observed data $Y(\bs)$ come from a skew-$t$ (Zhang and El-Shaarawi, 2012)
    \begin{align*}
      Y(\bs) = \bX(\bs)\bbeta + \lambda |z| + v(\bs)
    \end{align*}
    where
    \begin{itemize} \setlength{\itemsep}{0.25em}
      \item $\lambda \in \calR$ controls the skewness
      \item $z \sim N(0, \sigma^2)$ is a random effect
      \item $v(\bs)$ is a Gaussian process with variance $\sigma^2$ and \Matern correlation
      \item $\sigma^2 \sim \text{IG}(a, b)$
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Spatial skew-$t$ distribution}
  \begin{itemize} \setlength{\itemsep}{0.5em}
   \item \alert{Conditioned} on $z$ and $\sigma^2$, $Y(\bs)$ is a Gaussian spatial model
    \item Can use standard geostatistical methods to fit this model
    \item Predictions can be made through Kriging
  \end{itemize}
\end{frame}

\begin{frame}{Spatial skew-$t$ distribution}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item \alert{Marginalizing} over $z$ and $\sigma^2$ (via MCMC),
    \begin{align*}
      Y(\bs) \sim \text{skew-t}(\bX(\bs), \bOmega, \alpha, \text{df}=2a)
    \end{align*}
    where
    \begin{itemize}
      \item $\bX(\bs) \bbeta$ is the location
      \item $\bOmega = \frac{1}{ab} \bar{\bOmega}$ is a correlation matrix
      \item $\bar{\bOmega} = (\bSigma + \lambda^2 \bOne \bOne^T)$
      \item $\bSigma$ is a postive definite correlation matrix
      \item $\alpha = \lambda( 1 + \lambda^2 \bOne^T \bSigma^{-1} \bOne)^{-1/2} \bOne^T \bSigma^{-1}$ controls the skewness
    \end{itemize}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Extension of the skew-$t$ distribution}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Skew-$t$ distribution addresses two modeling concerns
    \begin{itemize}
      \item Extremal dependence
      \item Reasonable computing
    \end{itemize}
    \item Our contribution is to extend the skew-$t$
    \begin{itemize}
      \item Censoring to focus on extreme observations
      \item Partitioning to address long-range dependence
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Censoring data to focus on tail behavior}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item We censor the observed data at a high threshold $T$
    \item Censored data
    \begin{align*}
      \tilde{Y}_t(\bs) = \left\{ \begin{array}{ll}
          Y_t(\bs) \quad & \delta(\bs) = 1\\
          T \quad & \delta(\bs) = 0
      \end{array}\right.
    \end{align*}
    where $\delta(\bs) = I[Y(\bs) > T]$
    \item Allows tails of the distribution to speak for themselves
  \end{itemize}
\end{frame}

\begin{frame}{Spatial skew-$t$ distribution properties}
  \begin{itemize} \setlength{\itemsep}{0.5em}
      \item Model properties
    \begin{itemize}
    	\item Has flexible tail
      \begin{itemize}
        \item Skewness controlled by $\lambda$
        \item Weight of tails controlled by degrees of freedom $2a$
      \end{itemize}
    	\item For a skew-$t$ distribution $\lim_{c \rightarrow \infty} \chi(h) > 0$ (Padoan, 2011)
    	\item Computation that is on the order of Gaussian computation
    \end{itemize}
    \item Challenge: $\chi(h) > 0$ as $h \rightarrow \infty$ (Padoan, 2011)
    \begin{itemize}
      \item This occurs because all observations (near and far) share the same $z$ and $\sigma^2$
      \item We deal with this through a daily random partition (similar to Kim et al., 2005 for non-extreme modeling)
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Random partition}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Daily random partition allows $z$ and $\sigma^2$ to vary by site
    \begin{align*}
      Y(\bs) = \bX(\bs) \bbeta + \lambda z(\bs) + \sigma(\bs) v(\bs)
    \end{align*}
    \item Consider a set of knots $\bw_{k} \sim$ Uniform that define a random partition
    $P_{1}, \ldots, P_{K}$ such that
    \begin{align*}
      P_{k} = \{s : k = \argmin_\ell|| \bs - \bw_{\ell}|| \}
    \end{align*}
    where $\bw = (w_1, w_2)$
    \item For $\bs \in P_{k}$
    \begin{align*}
      z(\bs) &= z_{k}\\
      \sigma^2(\bs) &= \sigma^2_{k}
    \end{align*}
  \end{itemize}
\end{frame}

\begin{frame}{Random partition}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Within each partition $Y(\bs)$ has the same MV skew-t distribution as before
    \item Across partitions $Y(\bs)$ are asymptotically independent, but still correlated through $v(\bs)$
  \end{itemize}
\end{frame}

@@

@@

\begin{frame}{Random partition skew-$t$ model}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item This new model is called a random partition skew-$t$ model, and it has the properties we desire
    \begin{itemize}
      \item Marginal distribution with flexible tails
      \begin{itemize}
        \item $\lambda$ term allows for asymmetry
        \item Degrees of freedom control heavy vs light tails
      \end{itemize}
      \item Asymptotic spatial dependence for that decays with distance between sites through partitioning
      \item Computation is on the order of Gaussian models for large space-time datasets
    \end{itemize}
  \end{itemize}
\end{frame}

@@

\begin{frame}{MCMC details}
  \begin{itemize} \setlength{\itemsep}{0.5em}
    \item Three main steps
    \begin{enumerate}[1.]
      \item Impute censored data below $T$
      \item Update parameters with standard random walk Metropolis Hastings or Gibbs sampling
      \item Make spatial predictions
    \end{enumerate}
    \item Priors are selected to be conjugate when possible
  \end{itemize}
\end{frame}

>>>>

\begin{frame}{Motivation}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item Average behavior is important to understand, but it does not paint the whole picture
    \begin{itemize}
      \item e.g. When constructing river levees, engineers need to be able to estimate a 100-year or 1000-year flood levels
      \item e.g. Probability of exceeding a certain threshold level
    \end{itemize}
    \item Estimating the probability of rare events is challenging because these events are, by definition, rare
    \item Spatial extremes is promising because it borrows informaation across space
    \item Spatial extremes is also useful for estimating probability of extremes at sites without data
  \end{itemize}
\end{frame}

@@

\begin{frame}{Motivation}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item Ozone compliance for Clean Air Act (EPA)
    \begin{itemize}
      \item Annual fourth-highest daily maximum 8-hour concentration, averaged over 3 years, not to exceed 75 ppb
      \item Annual fourth-highest is the 99th percentile for the year
      \item Common objectives are
      \begin{itemize}
        \item To interpolate to unmonitored sites
        \item Detect changes in extremes over time
        \item Study factors that lead to extreme events
      \end{itemize}
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Defining extremes}
\begin{columns}[c]
\column{.45 \linewidth}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item Key in extreme value analysis is to define extremes
    \item Typically done in one of two ways
    \begin{itemize}
      \item Block maxima (red dots)
      \item Values over threshold considered extreme
    \end{itemize}
  \end{itemize}

  \column{.55\linewidth}
  \begin{figure}
    \includegraphics[width=1\linewidth, trim=0 0.5in 0 1in]{./plots/define_extreme.pdf}
    \caption{Hypothetical monthly maximums recorded over a three years}
    \end{figure}
\end{columns}
\end{frame}

\begin{frame}{Standard non-spatial analysis: Block maxima}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item Let $X_1, \ldots, X_n$ be i.i.d.
    \item Consider the block maximum $M_n = \max(X_1, \ldots, X_n)$
    \item If there exist normalizing sequences $a_n > 0$ and $b_n \in \calR$ such that
    \begin{align*}
      \frac{ M_n - b_n }{ a_n } \converged G(z)
    \end{align*}
    then $G(z)$ follows a generalized extreme value distribution (GEV) (Falk et al., 2011)
    \item This motivates the use of the GEV for block max data
  \end{itemize}
\end{frame}

\begin{frame}{Standard non-spatial analysis: Block maxima}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item GEV distribution
    \begin{align*}
      G(y) = \Pr(Y < y) = \left\{  \begin{array}{ll}
        \exp\left\{ -\left[ 1 + \xi \left( \frac{ y - \mu }{ \sigma } \right) \right]^{ -1 / \xi} \right\} & \quad \xi \neq 0 \\[0.5em]
        \exp \left\{ -\exp \left( - \frac{ y - \mu }{ \sigma} \right) \right\} & \quad \xi = 0
      \end{array}\right.
    \end{align*}
    where
    \begin{itemize}
      \item $\mu \in \calR$ is a location parameter
      \item $\sigma > 0$ is a scale parameter
      \item $\xi \in \calR$ is a shape parameter
      \begin{itemize}
        \item Unbounded above if $\xi \ge 0$
        \item Bounded above by $(\mu - \sigma) / \xi$ when $\xi < 0$
      \end{itemize}
    \end{itemize}
    \item Challenges:
    \begin{itemize}
      \item Lose information by only considering maximum in a block
      \item Underlying data may not be i.i.d.
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Standard non-spatial analysis: Peaks over threshold}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item Let $X_1, \ldots, X_n \sim F$
    \item If there exist normalizing sequences $a_t > 0$ and $b_t \in \calR$ such that for any $x \ge 0$, as $T \rightarrow \infty$
    \begin{align*}
      \Pr\left(\frac{X - b_t}{a_t} > t \mid X > T\right) \converged H(x),
    \end{align*}
    where $T$ is a thresholding value, then $H(x)$ follows a generalized Pareto distribution (GPD) (Balkema and de Haan, 1974)
  \end{itemize}
\end{frame}

\begin{frame}{Standard non-spatial analysis: Peaks over threshold}
  \begin{itemize}  \setlength{\itemsep}{1em}
    \item Select a threshold, $T$, and use the Generalized Pareto distribution to model the exceedances
    \begin{align*}
      H(y) = P(Y < y) = \left\{ \begin{array}{ll}
        1 - \left[1 - \xi \left( \frac{ y - T }{ \sigma } \right) \right]^{-1 / \xi} & \quad \xi \neq 0 \\[0.5em]
        1 - \exp \left\{ \frac{ y - T }{ \sigma} \right\} & \quad \xi = 0
      \end{array}\right.
    \end{align*}
    where
    \begin{itemize}
      \item $\sigma > 0$ is a scale parameter
      \item $\xi \in \calR$ is a shape parameter
      \begin{itemize}
        \item Unbounded above if $\xi \ge 0$
        \item Bounded above by $(T - \sigma) / \xi$ when $\xi < 0$
      \end{itemize}
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Standard non-spatial analysis: Peaks over threshold}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item The GPD is related to GEV distribution through
    \begin{align*}
      H(y) = 1 + \log[G(y)], \quad y \ge T
    \end{align*}
    \item Challenges:
    \begin{itemize}
      \item Sensitive to threshold selection
      \item Temporal dependence between observations (e.g. flood levels don't dissipate overnight)
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Max-stable processes for spatial data}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item Consider i.i.d. spatial processes $x_j(\bs)$, $j = 1, \ldots, J$
    \item Let $M_J(\bs) = \bigvee_{j=1}^J x_j(\bs_i)$ be the block max at site $\bs$
    \item If there exists normalizing sequences $a_J(\bs)$ and $b_J(\bs)$ such that for all sites, $\bs_i, i = 1, \ldots, d$,
    \begin{align*}
      a_J^{-1}(\bs) \left\{ M_J(\bs) - b_J(\bs) \right\} \converged Y(\bs)
    \end{align*}
    which has a non-degenerate distribution, then $Y(\bs)$ is a max-stable process
    \item Therefore, max-stable processes are the standard for block maxima
  \end{itemize}
\end{frame}

\begin{frame}{Multivariate representations}
  \begin{itemize}
    \item Marginally at each site, observations follow a GEV distribution
    \item For a finite collection of sites the multivariate representation for the GEV (mGEV) is
    \begin{align*}
      \Pr(\bZ \le \bz)  &= G^*(\bz) = \exp(-V(\bz))\\
            V(\bz)    &= d \int_{\Delta_d} \bigvee_{i = 1}^d \frac{w_i}{z_i} H(\ddd w)
    \end{align*}
    where
    \begin{itemize}
      \item $\Delta_d = \{ \bw \in \calR^d_+ \mid w_1 + \cdots + w_d = 1\}$
      \item $H$ is a probability measure on $\Delta_d$
      \item $\int_{\Delta_d}w_i H(\ddd w) = 1 / d$ for $i = 1, \ldots, d$.
    \end{itemize}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Multivariate peaks over threshold}
  \begin{itemize}
    \item Few existing methods
    \item Often use max-stable methods due to the relationship between GEV and GPD
    \item Joint distribution function given by Falk et al. (2011)
    \begin{align*}
      H(\bz) = 1 - V(\bz)
    \end{align*}
    where $V(\bz)$ is defined as in the GEV
  \end{itemize}
\end{frame}

\begin{frame}{Extremal dependence: $\chi$ statistic}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item Correlation is the most common measure of dependence
    \begin{itemize}
      \item Focuses on the center and not tails
      \item This makes it irrelevant for extreme value analysis
    \end{itemize}
    \item Extreme value analysis focuses on the $\chi$ statistic, a measure of extremal dependence given by
    \begin{align*}
      \chi(h) = \lim_{c \rightarrow \infty}\Pr[Y(\bs) > c \mid Y(\bt) > c]
    \end{align*}
    where $h = ||\bs - \bt||$
    \item If $ \chi(h) = 0$, then observations are asymptotically independent at distance $h$
  \end{itemize}
\end{frame}

\begin{frame}{Existing challenges}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item Multivariate max-stable and GPD models have nice features, but they are
    \begin{itemize}
      \item Computationally challenging (e.g,  the asymmetric logistic has $2^{n-1}(n + 2) - (2n + 1)$ free parameters)
      \item Joint distribution only available in low dimensions
    \end{itemize}
    \item Some recent approaches
    \begin{itemize}
      \item Bayesian hierarchical model (Reich and Shaby, 2012)
      \item Pairwise likelihood approach (Huser and Davison, 2014)
    \end{itemize}
    \item Many opportunities to explore new methods
  \end{itemize}
\end{frame}

\begin{frame}{Three principal contributions}
  \begin{enumerate}[1] \setlength{\itemsep}{1em}
    \item A spatio-temporal model with flexible tails, asymptotic spatial dependence, and computation on the order of Gaussian models for large space-time datasets
    \item Predicting rare binary events with a spatially dependent generalized extreme-value link function
    \item A Bayesian hierarchical model to allow for non-stationary covariance in extreme value models
  \end{enumerate}
\end{frame}

\begin{frame}{Spatiotemporal modeling for extreme values}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item Model objectives:
    \begin{itemize}
      \item Marginal distribution with a flexible tail
      \begin{itemize}
        \item Allow for asymmetric distributions
        \item Allow for heavy tails
      \end{itemize}
      \item Asymptotic spatial dependence
      \item Computation on the order of Gaussian models for large space-time datasets
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Gaussian spatial model}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item In geostatistics $Y(\bs)$ are often modeled using a Gaussian process with mean function $\mu(\bs)$ and covariance function $\rho(h)$.
    \item Model properties
    \begin{itemize}
      \item Nice computing properties (closed-form likelihood)
      \item For a Gaussian spatial model $\lim_{c \rightarrow \infty} \chi(h) = 0$ regardless of the strength of the correlation in the bulk of the distribution
      \item Tail is not flexible
      \begin{itemize}
        \item Light-tailed
        \item Symmetric
      \end{itemize}
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Spatial skew-$t$ distribution}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item A more flexible alternative is the spatial skew-$t$ process (Zhang and El-Shaarawi, 2012)
    \begin{align*}
      Y(\bs) = \bX(\bs)\bbeta + \lambda |z| + v(\bs)
    \end{align*}
    where
    \begin{itemize} \setlength{\itemsep}{0.25em}
      \item $\lambda \in \calR$ controls the skewness
      \item $z \sim N(0, \sigma^2)$ is a random effect
      \item $v(\bs)$ is a Gaussian process with variance $\sigma^2$ and \Matern correlation
      \item $\sigma^2 \sim \text{IG}(a, b)$
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Spatial skew-$t$ distribution}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item \alert{Conditioned} on $z$ and $\sigma^2$, $Y(\bs)$ is a Gaussian spatial model
    \item Can use standard geostatistical methods to fit this model
    \item Predictions can be made through Kriging
  \end{itemize}
\end{frame}

\begin{frame}{Spatial skew-$t$ distribution}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item \alert{Marginalizing} over $z$ and $\sigma^2$ (via MCMC),
    \begin{align*}
      Y(\bs) \sim \text{skew-t}(\bX(\bs), \bOmega, \alpha, \text{df}=2a)
    \end{align*}
    where
    \begin{itemize}
      \item $\bX(\bs) \bbeta$ is the location
      \item $\bOmega = \frac{1}{ab} \bar{\bOmega}$ is a correlation matrix
      \item $\bar{\bOmega} = (\bSigma + \lambda^2 \bOne \bOne^T)$
      \item $\bSigma$ is a postive definite correlation matrix
      \item $\alpha = \lambda( 1 + \lambda^2 \bOne^T \bSigma^{-1} \bOne)^{-1/2} \bOne^T \bSigma^{-1}$ controls the skewness
    \end{itemize}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Spatial skew-$t$ distribution properties}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item Model properties:
    \begin{itemize}
      \item Flexible tail
      \begin{itemize}
        \item Skewness controlled by $\lambda$
        \item Weight of tails controlled by degrees of freedom $2a$
      \end{itemize}
      \item For a skew-$t$ distribution $\lim_{c \rightarrow \infty} \chi(h) > 0$ (Padoan, 2011)
      \item Computation that is on the order of Gaussian computation
    \end{itemize}
    \item Challenge: $\chi(h) > 0$ as $h \rightarrow \infty$ (Padoan, 2011)
    \begin{itemize}
      \item This occurs because all observations (near and far) share the same $z$ and $\sigma^2$
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Extension of the skew-$t$ distribution}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item Skew-$t$ distribution addresses two modeling concerns
    \begin{itemize}
      \item Extremal dependence
      \item Reasonable computing
    \end{itemize}
    \item Our contribution is to extend the skew-$t$
    \begin{itemize}
      \item Censoring to focus on extreme observations
      \item Partitioning to address long-range dependence
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Censoring data to focus on tail behavior}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item We censor the observed data at a high threshold $T$
    \item Censored data
    \begin{align*}
      \tilde{Y}_t(\bs) = \left\{ \begin{array}{ll}
          Y_t(\bs) \quad & \delta(\bs) = 1\\
          T \quad & \delta(\bs) = 0
      \end{array}\right.
    \end{align*}
    where $\delta(\bs) = I[Y(\bs) > T]$
    \item Allows tails of the distribution to speak for themselves
  \end{itemize}
\end{frame}

\begin{frame}{Random partition}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item Daily random partition allows $z$ and $\sigma^2$ to vary by site
    \begin{align*}
      Y(\bs) = \bX(\bs) \bbeta + \lambda z(\bs) + \sigma(\bs) v(\bs)
    \end{align*}
    \item Consider a set of knots $\bw_{k} \sim$ Uniform that define a random partition
    $P_{1}, \ldots, P_{K}$ such that
    \begin{align*}
      P_{k} = \{s : k = \argmin_\ell|| \bs - \bw_{\ell}|| \}
    \end{align*}
    where $\bw = (w_1, w_2)$ (similar to Kim et al., 2005 for non-extreme modeling)
    \item For $\bs \in P_{k}$
    \begin{align*}
      z(\bs) &= z_{k}\\
      \sigma^2(\bs) &= \sigma^2_{k}
    \end{align*}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Random partition}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item Within each partition $Y(\bs)$ has the same MV skew-t distribution as before
    \item Across partitions $Y(\bs)$ are asymptotically independent, but still correlated through $v(\bs)$
    \item New expression for $\chi(h)$
    \begin{align*}
      \chi(h) = \pi(h) \chi_{\text{skew-}t}(h)
    \end{align*}
    where
    \begin{itemize}
      \item $\pi(h)$ is the probability two sites are in the same partition
    \end{itemize}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Random partition skew-$t$ model}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item This new model is called a random partition skew-$t$ model, and it has the properties we desire
    \begin{itemize}
      \item Marginal distribution with flexible tails
      \begin{itemize}
        \item $\lambda$ term allows for asymmetry
        \item Degrees of freedom control heavy vs light tails
      \end{itemize}
      \item Asymptotic spatial dependence for that decays with distance between sites through partitioning
      \item Computation is on the order of Gaussian models for large space-time datasets
    \end{itemize}
  \end{itemize}
\end{frame}

@@

\begin{frame}{MCMC details}
  \begin{itemize} \setlength{\itemsep}{1em}
    \item Three main steps
    \begin{enumerate}[1.]
      \item Impute censored data below $T$
      \item Update parameters with standard random walk Metropolis Hastings or Gibbs sampling
      \item Make spatial predictions
    \end{enumerate}
    \item Priors are selected to be conjugate when possible
  \end{itemize}
\end{frame}
